High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-23 doc/asplos2013/interface.tex

\section{Programming Interface}
\label{sec:interface}
The target set of clients for our interface are higher-level language
compilers and runtimes such as Legion\cite{Legion12} as well as 
advanced systems programmers.  This class of users demand total control
over the underlying hardware and transparent performance from an
implementation.  To support these demands we have designed our interface
to be as low-level as possible.  In many cases we have chosen to trade-off
programmability for performance.  We eschewed any features that would
automatically be performed without being directed by the client.
Our goal in the design of this interface is to provide a set of low-level,
high-performance primitives that are asynchronous and composable.

@@

In most cases events are created as the result of other
operations and the implementation is responsible for triggering these events.  Users 
can create a special type of event called a {\tt UserEvent} (line 15) that the
user has the power to trigger.  A user event is a sub-type of an event
that has a {\tt trigger} method (line 18).  Users can also create another
type of event called a {\tt Barrier} that will require multiple arrivals before
the event is triggered.  The user can manage the number of arrivals that must
be seen as well as how many arrivals occur at a time.  Once a barrier's arrival
count goes to zero, the barrier will trigger.  Note that barriers in our interface
provide a superset of the functionality of traditional barriers since they can
be used in a blocking manner (via the {\tt wait} call), but are primarily used
asynchronously.  Lastly, since user events and barriers are sub-types of 
an event they can be used wherever an event is required.

@@

class Barrier : public Event {
public:
  static Barrier create_barrier(unsigned expected_arrivals);
  void alter_arrival_count(int delta) const;
  void arrive(unsigned count = 1) const;
};
\end{lstlisting}

\subsection{Processors}
\label{subsec:procs}
Listing~\ref{lst:procapi} shows the interface for {\tt Processor} objects which allow for
the creation of parallel computations.  Processor provide a way of naming 
every computational unit in the machine.  We describe how to discover processor handles
in Section~\ref{subsec:machmodel}.  In the case of our current interface 
processors are either individual cores or discrete GPUs, but this is easily modified 
to include new processor types (line 10).   Processors support a single {\tt spawn}
operation (lines 13-14) that will launch a new {\em task} on that processor.
Note that because the spawn operation
is invoked on a processor handle, it is possible to launch a task on any
processor from anywhere in the system.  
The spawn call takes an optional event that must trigger before the task can begin.  
The spawn call returns an event that will trigger when the task completes.

Lines 17-23 of Listing~\ref{lst:procapi} show a simple example using 
the processor and event interface.  The {\tt diamond\_task} launches four
sub-tasks with a diamond dependence pattern.  The {\tt merge\_events} call
is used to merge the events {\tt eb} and {\tt ec} to create the shared dependence
for the last sub-task.

@@

  enum Kind {
    CPU_PROC,GPU_PROC,// ... future processor types
  };

@@

Between deffered locks and barriers (described in Section~\ref{subsec:events}) clients 
have access to the same set of synchronization primitves that they traditionally have
in both threading and bulk-synchronous interfaces.  However, these operations can now 
be composed with other asynchronous operations which is not possible in any other interface
of which we are aware.

@@

A physical region is an allocation of data in a single memory in the memory hierarchy.  Physical
regions are grouped into {\em classes} which are sets of physical regions that share the
same number and size of elements, but don't necessarily share the same data layout.  Our
interface supports sub-classing, but the details are omitted due to space constraints.  The
interface only supports data movement operations between physical regions that are of
the same class (or between super- and sub-classes).  Note, clients can still pass arbitrary arrays
of bits via the processor {\tt spawn} call.
% Say something about dynamically modifying the number of elements in a class

@@

Reductions and reduction instances illustrate only a single case where physical regions provide 
the structure necessary to conjoin an operation with data movement.  Physcal regions make our
interface easily extensible to arbitrary transformations and operations as a part of data movment.
While physical regions are a slightly higher-level construct than an annonymous
buffer of bits, we believe that the optimizations they enable are important in an asynchronous
environment.  Clients also still have access to an interface for controlling data layout through
the operations associated with physical regions which mitigates the loss in generality.

@@

\subsection{Machine}
\label{subsec:machmodel}
The last component of the interface allows for introspection of the current machine.
Listing~\ref{lst:machineapi} shows the machine interface.
The {\tt Machine} object is a singleton object that serves two purposes.  First, it
is the mechanism for initializing the runtime; at the start of a program the client 
creates a machine object, instantiates a task table mapping task IDs to function pointers,
and invokes the {\tt run} method (line 30-31).
Second, the machine object provides an interface for the client to 
discover properties of the underlying machine (lines 35-43).  The machine object maintains
sets of all the unique {\tt Processor} and {\tt Memory} handles in the system.  Processors
were described in Section~\ref{subsec:procs}.  For every memory in the system there also 
exists a {\tt Memory} handle with a unique ID.  Different memories often, but not always, 
imply a different address space.  Examples of different memories are described in
Section~\ref{sec:impl}.

@@

The machine object also maintains two relations between the set of processors
and the set of memories:
\begin{itemize}
\item Processor-Memory: for each processor-memory pair whether
the processor can directly access (read and write) the memory and 
latency and bandwidth properties if access is possible
\item Memory-Memory: for every pair of memories whether copies can be
performed and latency and bandwidth properties if copies are possible
\end{itemize}

The client can query these two relations by the {\tt get\_*\_affinity} calls
(lines 40-43) to discover properties of the machine.

@@

\begin{lstlisting}[float={t},label={lst:machineapi},caption={Machine Interface.}]

class Memory {
public:
  const unsigned id;
};

class Machine {
public:
  Machine(int *argc, char ***argv,
          const Processor::TaskIDTable &task_table);
  enum RunStyle {
    ONE_TASK_ONLY,ONE_TASK_PER_NODE,
    ONE_TASK_PER_PROCESSOR,
  };
  void run(Processor::TaskFuncID task_id = 0, 
        RunStyle = ONE_TASK_ONLY, const void *args, size_t arglen);
public:
  static Machine* get_machine(void); // pointer to Machine
  // References to all processors and memories
  const set<Memory>& get_all_memories(void) const;
  const set<Processor>& get_all_processors(void) const;
  Processor::Kind get_processor_kind(Processor p) const;
  size_t get_memory_size(Memory m) const;
  // Query properties of underlying hardware
  int get_proc_mem_affinity(vector<ProcMemAffinity &result,
          Processor restrict_proc = 0, Memory restrict_mem = 0);
  int get_mem_mem_affinity(vector<MemMemAffinity> &result,
          Memory restrict_mem1 = 0, Memory restrict_mem2 = 0);
};
\end{lstlisting}

>>>>

\section{Programming Interface}
\label{sec:interface}
The target set of clients for our interface are higher-level language
compilers and runtimes such as Legion\cite{Legion12} as well as 
advanced systems programmers.  This class of users demand total control
over the underlying hardware and transparent performance from an
implementation.  To support these demands we have designed our interface
to be as low-level as possible.  In many cases we have chosen to trade-off
ease of programmability for performance.  We eschewed any features that would
automatically be performed without being directed by the client.
Our goal in the design of this interface is to provide a set of low-level,
high-performance primitives that are asynchronous and composable.

@@

In most cases events are created as the result of other
operations and the implementation is responsible for triggering these events.  Clients
can also create a {\tt UserEvent} (line 15) that are triggered explicitly by the client.

The common synchronization pattern in which multiple consumers are dependent on a 
(potentially different) set of producers is supported by a {\tt Barrier} object,
which is similar to a {\tt UserEvent}, but does not trigger until the expected
number of {\tt arrive} operations have occurred.  Barrier constructs exist in other
programming models\cite{MPI}, but in addition to the decoupling of the waiters from
the arrivers, our barrier interface has two further improvements to better support
composability and asynchronous operation:
\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
\item An arrival operation can be made dependent on another event, allowing a launch 
of a subtask or a copy of data to be completed before the barrier can trigger.
\item The expected arrival count of a barrier can be dynamically increased, allowing
a producer task to be split into multiple subtasks without the original task having to wait for
their completion to signal its arrival.
\end{itemize}

Although {\tt UserEvent}s and {\tt Barrier}s are created and triggered differently,
any operation that expresses a dependence on an {\tt Event} can transparently use
a {\tt UserEvent} or {\tt Barrier} as well.

%%   Users can also create another
%% type of event called a {\tt Barrier} that will require multiple arrivals before
%% the event is triggered.  The user can manage the number of arrivals that must
%% be seen as well as how many arrivals occur at a time.  Once a barrier's arrival
%% count goes to zero, the barrier will trigger.  Note that barriers in our interface
%% provide a superset of the functionality of traditional barriers since they can
%% be used in a blocking manner (via the {\tt wait} call), but are primarily used
%% asynchronously.  Lastly, since user events and barriers are sub-types of 
%% an event they can be used wherever an event is required.

@@

class Barrier : public Event {
public:
  static Barrier create_barrier(unsigned expected_arrivals);
  void alter_arrival_count(int delta) const;
  void arrive(unsigned count = 1,
                  Event wait_for = Event::NO_EVENT) const;
};
\end{lstlisting}

\subsection{Processors}
\label{subsec:procs}
Listing~\ref{lst:procapi} shows the interface for {\tt Processor} objects which allow for
the creation of parallel computations.  Processor provide a way of naming 
every computational unit in the machine.  We describe how to discover processor handles
in Section~\ref{subsec:machmodel}.  In the case of our current interface 
processors are either individual cores or discrete GPUs, but this is easily modified 
to include new processor types (line 10).   Processors support a single {\tt spawn}
operation (lines 14-15) that will launch a new {\em task} on that processor.
Note that because the spawn operation
is invoked on a processor handle, it is possible to launch a task on any
processor from anywhere in the system.  
The spawn call takes an optional event that must trigger before the task can begin.  
The spawn call returns an event that will trigger when the task completes.

Lines 18-24 of Listing~\ref{lst:procapi} show a simple example using 
the processor and event interface.  The {\tt diamond\_task} launches four
sub-tasks with a diamond dependence pattern.  The {\tt merge\_events} call
is used to merge the events {\tt eb} and {\tt ec} to create the shared dependence
for the last sub-task.

@@

  enum Kind {
    CPU_PROC,GPU_PROC,// ... future processor types
  };
  Kind kind() const;

@@

Between deferred locks and barriers (described in Section~\ref{subsec:events}) clients 
have access to the same set of synchronization primitves that they traditionally have
in both threading and bulk-synchronous interfaces.  However, these operations can now 
be composed with other asynchronous operations which is not possible in any other interface
of which we are aware.

@@

A physical region is an allocation of data in a single memory in the memory hierarchy.  Physical
regions are grouped into {\em classes} which are sets of physical regions that share the
same names (i.e. addresses) for elements, but don't necessarily share the same data layout.  This
allows pointers to elements within regions to be used, even if the regions are copied from one memory
to another.  Our
interface supports sub-classing, but the details are omitted due to space constraints.  The
interface allows copies between any pair of physical regions that are of
the same class (or between super- and sub-classes).

%Note, clients can still pass arbitrary arrays
%of bits via the processor {\tt spawn} call.

% Say something about dynamically modifying the number of elements in a class

@@

%% SJT - redundant with first paragraph in this subsection?
%% Reductions and reduction instances illustrate only a single case where physical regions provide 
%% the structure necessary to conjoin an operation with data movement.  Physcal regions make our
%% interface easily extensible to arbitrary transformations and operations as a part of data movment.
%% While physical regions are a slightly higher-level construct than an annonymous
%% buffer of bits, we believe that the optimizations they enable are important in an asynchronous
%% environment.  Clients also still have access to an interface for controlling data layout through
%% the operations associated with physical regions which mitigates the loss in generality.

@@

\subsection{Machine}
\label{subsec:machmodel}
The last component of the interface is shown in Listing~\ref{lst:machineapi} and allows the
initialization and run-time introspection
of the current machine.
The {\tt Machine} object is a singleton object that the client creates at the start of a
program, providing a task table that maps task IDs to function pointers,
and then invoking the {\tt run} method (line 15-16).
Any running task can get a pointer to the {\tt Machine} object with the
{\tt get\_machine} method (line 18) and use it to obtain lists of all the {\tt Memory} (line 19) or 
{\tt Processor} (line 20) handles in the system.  Using the affinity methods (lines 22-25),
the application can also determine:
\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
\item for each {\tt Processor}, which {\tt Memory} objects are accessible and with what performance, and
\item which pairs of {\tt Memory} objects are able to efficiently copy data between them.
\end{itemize}

%% in the machine.
%% Second, the machine object provides an interface for the client to 
%% discover properties of the underlying machine (lines 35-43).  The machine object maintains
%% sets of all the unique {\tt Processor} and {\tt Memory} handles in the system.  Processors
%% were described in Section~\ref{subsec:procs}.  For every memory in the system there also 
%% exists a {\tt Memory} handle with a unique ID.  Different memories often, but not always, 
%% imply a different address space.  Examples of different memories are described in
%% Section~\ref{sec:impl}.

@@

%% The machine object also maintains two relations between the set of processors
%% and the set of memories:
%% \begin{itemize}
%% \item Processor-Memory: for each processor-memory pair whether
%% the processor can directly access (read and write) the memory and 
%% latency and bandwidth properties if access is possible
%% \item Memory-Memory: for every pair of memories whether copies can be
%% performed and latency and bandwidth properties if copies are possible
%% \end{itemize}
%%
%% The client can query these two relations by the {\tt get\_*\_affinity} calls
%% (lines 40-43) to discover properties of the machine.

@@

\begin{lstlisting}[float={t},label={lst:machineapi},caption={Machine Interface.}]
class Memory {
public:
  const unsigned id;
  size_t size() const;
};

class Machine {
public:
  Machine(int *argc, char ***argv,
          const Processor::TaskIDTable &task_table);
  enum RunStyle {
    ONE_TASK_ONLY,ONE_TASK_PER_NODE,
    ONE_TASK_PER_PROCESSOR,
  };
  void run(Processor::TaskFuncID task_id = 0, 
        RunStyle = ONE_TASK_ONLY, const void *args, size_t arglen);
public:
  static Machine* get_machine(void);
  const set<Memory>& get_all_memories(void) const;
  const set<Processor>& get_all_processors(void) const;

  int get_proc_mem_affinity(vector<ProcMemAffinity &result,
          Processor restrict_proc = 0, Memory restrict_mem = 0);
  int get_mem_mem_affinity(vector<MemMemAffinity> &result,
          Memory restrict_mem1 = 0, Memory restrict_mem2 = 0);
};
\end{lstlisting}

### 2012-07-23 doc/asplos2013/legion-llr.tex

\begin{abstract}
%The limit is 12 pages, including bibliography and appendices.
%Deadlines:
%\begin{tabular}{ll}
%Abstract deadline & July 16, 2012 3pm PDT \\
%Paper deadline & July 23, 2012 3pm PDT \\
%Author response & around October 15, 2012 \\
%Author notification & November 9, 2012
%\end{tabular}
We present a low-level runtime interface for current highly distributed, heterogeneous
parallel machines that provides a fully asynchronous 
programming model, allowing all types of runtime actions, including synchronization,
to be non-blocking.  Our interface includes mechanisms for the creation and movement
of data through the memory hierarchy as well as {\em deferred locks}, a novel synchronization primitive.
Asynchrony is exposed by the the interface using a distributed event system.  All
operations return an event to be triggered when the operation completes.  Events
can be chained together and used to create dependences between operations.

>>>>

\begin{abstract}
%The limit is 12 pages, including bibliography and appendices.
%Deadlines:
%\begin{tabular}{ll}
%Abstract deadline & July 16, 2012 3pm PDT \\
%Paper deadline & July 23, 2012 3pm PDT \\
%Author response & around October 15, 2012 \\
%Author notification & November 9, 2012
%\end{tabular}
We present a low-level runtime interface for current highly distributed, heterogeneous
parallel machines that provides a fully asynchronous 
programming model, allowing all types of runtime actions, including synchronization,
to be non-blocking.  Our interface includes mechanisms for the creation and movement
of data through the memory hierarchy as well as {\em deferred locks}, a novel synchronization primitive.
Asynchrony is exposed by the interface using a distributed event system.  All
operations return an event to be triggered when the operation completes.  Events
can be chained together and used to capture dependences between operations.

### 2012-07-23 doc/asplos2013/results.tex

\subsection{Event Lifetimes}
\label{subsec:eventlife}
We instrumented the heterogenous implementation of our interface to 
capture information about different kinds of events.  Figure~\ref{fig:eventlife} shows
the lifetimes of different types of events from a single run of the Fluid
application on 16 nodes using 8 processors per node.  Dynamic events 
is a monotonically increasing line that
corresponds to the total number of events created in the system.  By
the end of the run, the simulation created over 260,000 dynamic
events.  In contrast, the physical events line corresponds to the total number
of generational events required to represent the dynamic events 
across all the nodes in the system.  Less than 5,000
generational events were required to represent all the dynamic events
illustrating that the technique of mapping from dynamic events to generational 
events described in Section~\ref{subsec:eventimpl} was effective.

The live event line shows the total number of {\em live events} at a point in time.  Dynamic events
are considered {\em live events} until their last query has been performed.  The number of live
events is equivalent to the number of physical events that would be required
in a reference counted scheme.  In this case we see that by reusing
generational events, we require up to 4X fewer generational events than would be
required by a reference counted scheme.

In an ideal world the total number of generational events required would be
equivalent to the maximum number of generational events in the untriggered
state at any point.  The untriggered event line shows the number of generational
events in the untriggered state.  We see that the total number of generational
events is actually slightly higher than the maximum number of generational
events in the untriggered state at any point.  The reason for this is that unused
generational events cannot be shared across nodes, so in some cases nodes must
make new generational events eventhough there are unused generational events
on other nodes.  The small difference in these lines shows that in practice
this is only a minor inefficiency.

>>>>

\subsection{Event Lifetimes}
\label{subsec:eventlife}

We instrumented the heterogenous implementation of our interface to 
capture event usage information.  Figure~\ref{fig:eventlife} shows
a timeline of the execution of the Fluid application on 16 nodes.
The dynamic events line measures the total number of event creations.
A large number of events are created - over 260,000 in less than 15
seconds - and allocating separate storage for every event would clearly
be difficult for long-running applications.  The {\em live events} line
measures the number of events that are live at any given time - after
the last operation on an event, a reference counting implementation would
be able to recover the storage associated with the event.  Such an
algorithm would reduce the storage needed by over 10X at the cost of
computation and communication overhead for performing the reference 
counting.  As discussed in Section~\ref{subsec:eventimpl}, our implementation
requires storage that grows with the number of {\em untriggered events}, a number
that is 10X smaller than even the maximal live event count.  The actual storage
requirements of our implementation are shown by the {\em physical events} line,
which is larger than the peak number of untriggered events because of the
decision to always create an event owned by the node from which the request
was made.  Despite this inefficiency, our implementation uses 5X less storage
than a reference counting implementation and avoids any overhead related to 
reference counting.  These savings would likely be even more dramatic for longer 
runs of the application, as the number of live events is steadily growing as the
application runs, while the peak number of generational events needed appears to 
occur during the startup of the application.

%% much storage would be required for events using

%%  about different kinds of events.  Figure~\ref{fig:eventlife} shows
%% the lifetimes of different types of events from a single run of the Fluid
%% application on 16 nodes using 8 processors per node.  Dynamic events 
%% is a monotonically increasing line that
%% corresponds to the total number of events created in the system.  By
%% the end of the run, the simulation created over 260,000 dynamic
%% events.  In contrast, the physical events line corresponds to the total number
%% of generational events required to represent the dynamic events 
%% across all the nodes in the system.  Less than 5,000
%% generational events were required to represent all the dynamic events
%% illustrating that the technique of mapping from dynamic events to generational 
%% events described in Section~\ref{subsec:eventimpl} was effective.

%% The live event line shows the total number of {\em live events} at a point in time.  Dynamic events
%% are considered {\em live events} until their last query has been performed.  The number of live
%% events is equivalent to the number of physical events that would be required
%% in a reference counted scheme.  In this case we see that by reusing
%% generational events, we require up to 4X fewer generational events than would be
%% required by a reference counted scheme.

%% In an ideal world the total number of generational events required would be
%% equivalent to the maximum number of generational events in the untriggered
%% state at any point.  The untriggered event line shows the number of generational
%% events in the untriggered state.  We see that the total number of generational
%% events is actually slightly higher than the maximum number of generational
%% events in the untriggered state at any point.  The reason for this is that unused
%% generational events cannot be shared across nodes, so in some cases nodes must
%% make new generational events eventhough there are unused generational events
%% on other nodes.  The small difference in these lines shows that in practice
%% this is only a minor inefficiency.
